---
title: "Brookings Data Network: String Basics"
author: "Sifan Liu"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    nature:
      ratio: 16:9
---

# Hello!

```{r setup, include = F}
library(xaringanExtra)
xaringanExtra::use_panelset()
knitr::opts_chunk$set(warning=FALSE)
```

### What we will cover:
- What is a string?
- What can we do with strings in R?

--

### What we won't cover:
- Regular Expressions (RegEx)

???
My name is Sifan Liu, Senior Research Analyst at Brookings Metro. I am also the current co-chair of Brookings Data Network. 
Today, we are going to talk about how to work with strings in R. We will focus on the basics, without going into Regular Expressions.

---
# String Basics
- They are also known as "text", "characters", etc. 

--

- Strings are always quoted with single quotes `''` or double quotes `""`.
```{r define}
a <- "I am a string"
a
typeof(a)
```
--
```{r}
b <- c("We are", "also strings")
b
typeof(b)
```

```{r}
as.data.frame(b)
```

---
# String basics

.panelset[
.panel[.panel-name[Quiz]

Which of the following object is not a string?

1. TRUE

1. "16"

1. c("a", "30")

1. "\n"


]

.panel[.panel-name[Answer]

```{r error=TRUE}

typeof(TRUE)
typeof("16")
typeof(c("a", 30))
typeof("\n")
```
]
]

---

# Let's clean some strings!

```{r, include = F}
df <- tibble::tribble(
        ~state, ~state_code, ~state_name, ~county_code,          ~county,
          "AL",          1L,   "Alabama ",           1L, "Autauga County",
          "AL",          1L,   "Alabama",           3L, "Baldwin county",
          "AL",          1L,   " Alabama",           5L, "Barbour County",
          "AL",          1L,   "Alabama ",           7L,    "Bibb County",
          "AL",          1L,   "Alabama",           9L,  "Blount COUNTY",
          "AL",          1L,   "Alabama",          11L, "Bullock County"
        )

```

```{r}
df
```
--
```{r length, warning=FALSE}
library(stringr)
str_length(df$state_name)
```
---
# Managing length
### str_trim()
removes whitespace from a string
```{r}
df$state_name %>% 
  str_trim() %>%
  str_length()
```
--
### str_pad()
pads a string to a new length
```{r}
str_pad(df$county_code, width = 3, side = "left", pad = "0")
```

---
# Convert case

```{r}
str_to_lower("Blount COUNTY")
str_to_upper("Blount COUNTY")
str_to_title("Blount COUNTY")
```
---
# Subset and combine
## str_sub()
```{r}
str_sub(df$county, start = 1, end = -8)

```

--
## str_c()
```{r}
str_c(df$state, ",US")
```

---
# Detect, extract, replace
## str_detect()
```{r}
str_detect(df$county, "a")

df %>% 
  dplyr::filter(str_detect(county, "a"))
```

---
# Detect, extract, replace
## str_extract
```{r}
str_extract(df$county, "county")
```
## str_replace
```{r}
df$county %>% 
  str_to_lower() %>% 
  str_replace("county", "cty")
```

---
# Let's code!

---
# What's next?
## Advanced 
[stringi](https://stringi.gagolewski.com/)

--

## Regular Expressions
- define more complex search patterns
1. extract all numbers in the strings
2. remove all characters after the last `.`
3. find the string that ends with "s"

--

## Special forms of strings
- Dates and time: [lubridate](https://lubridate.tidyverse.org/)
- Human names: [humaniformat](https://cran.r-project.org/web/packages/humaniformat/vignettes/Introduction.html)
- Addresses: [tidygeocoder](https://cran.r-project.org/web/packages/tidygeocoder/vignettes/tidygeocoder.html)